Complementing theory on adolescent emotion regulation using  machine learning

### 2022-03-14 manuscript/manuscript.Rmd
Incorporate Lucas' comments

authornote: |
  CVL conceived of the study, conducted the final analyses, and wrote the draft paper. LB conducted preliminary analyses and gave feedback on the draft. SB and WM provided data and gave feedback on the draft. This work was funded by a NWO Veni Grant (NWO Grant Number VI.Veni.191G.090), awarded to the lead author. Data of the RADAR (Research on Adolescent Development And Relationships) study were used (https://doi.org/10.17026/dans-zrb-v5wp). RADAR has been financially supported by main grants from the Netherlands Organisation for Scientific Research (GB-MAGW 480-03-005, GB-MAGW 480-08-006), Stichting Achmea Slachtoffer en Samenleving (SASS), the Netherlands Organisation for Scientific Research to the Consortium Individual Development (CID; 024.001.003), a grant of the European Research Council (ERC-2017-CoG - 773023 INTRANSITION), and various other grants from the Netherlands Organisation for Scientific Research, VU University Amsterdam, and Utrecht University. The funders had no role in the study design, data collection and analysis, decision to publish, or preparation of the manuscript. If the regulations of the journal allow it, the participating children, parents and schools are to be thanked as well.

abstract: |
  This paper set out to identify the most important early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM forests. Candidate predictors were selected based on a prior text mining systematic review of 6305 papers. Participants were 497 Dutch adolescents. Predictors were assessed at age 13, and emotion regulation development from age 14-18 was described using a quadratic latent trajectory model. SEM-forests were used to group adolescents with similar trajectories, based on their predictor scores. This person-centered approach not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents may follow similar trajectories. The results indicated that individual differences (e.g., in personality) and conflict behaviors in relationships with parents and friends, and internalizing and externalizing were important early predictors. Less important were parenting behaviors, except autonomy support and invasive parenting, bullying and victimization, delinquent peers, and substance use. The implications of these findings for theory formation and intervention are discussed, and we present an open source risk assessment tool, ERRATA.

@@

Emotion regulation,
defined as the ability to manage emotional experience and expression,
is critical for mental health [@aldaoEmotionregulationStrategiesPsychopathology2010, @schaferEmotionRegulationStrategies2017, @braetEmotionRegulationChildren2014]
and social competence [@zimmermannEmotionRegulationEarly2014] throughout life.
There is mounting evidence, however, that adolescents do not uniformly become better at regulating emotions: 
staggered development of emotional and regulatory brain circuits gives rise to a "maturity gap" in mid-adolescence [@croneUnderstandingAdolescencePeriod2012].
The emotional salience of social stimuli is intensified, while the ability to regulate these emotions lags behind [@caseyBrakingAcceleratingAdolescent2011, @vanlissaLongitudinalInterplayAffective2014].
This imbalance motivates most adolescents to engage in adaptive exploration:
Seeking out experiences required to develop mature socio-emotional competences.
At the same time, this imbalance renders some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014].
<!-- exposes youth to a risk of emotion dysregulation, -->
For example, they might not yet be sufficiently equipped to regulate the pursuit of unhealthy incentives,
or persevere in pursuing adaptive goals when facing obstacles.
Thus, some become trapped in a negative spiral characterized by more negative and volatile emotions [@craccoEmotionRegulationChildhood2017],
which has been linked to the emergence of several mental disorders [@schaferEmotionRegulationStrategies2017].
<!-- This imbalance motivates most youngsters to seek out new challenges in life and love, -->
<!-- but renders some of them vulnerable to emotion dysregulation1,10. Unfortunately, there is little theory about which risk factors and environmental hazards render adolescents susceptible to emotional difficulties11. -->
<!-- Acquiring adaptive emotion regulation skills in adolescence plays a key role in this regard7,20, because t -->

@@

As we set out to identifying important predictors of emotion regulation development,
this latent class analysis is not sufficient to answer our research question.
In latent class analysis, individuals are clustered only with respect the the outcome (growth trajectories), to ensure maximum homogeneity of growth trajectories in each class.
It thus answers the question: "Which adolescents have similar growth trajectories?"
SEM-forests similarly aim to ensure maximum homogeneity of growth trajectories within clusters,
but they differ from latent class analysis in that they cluster individuals by reference to a set of predictors.
They thus answer the question: "Why do some adolescents have similar growth trajectories?"
This method is suitable to meet the study goals.

>>>>

authornote: |
  This is a preprint paper, generated from Git Commit # `r gert::git_commit_id()`. CVL conceived of the study, conducted the final analyses, and wrote the draft paper. LB conducted preliminary analyses and gave feedback on the draft. SB and WM provided data and gave feedback on the draft. This work was funded by a NWO Veni Grant (NWO Grant Number VI.Veni.191G.090), awarded to the lead author. Data of the RADAR (Research on Adolescent Development And Relationships) study were used (https://doi.org/10.17026/dans-zrb-v5wp). RADAR has been financially supported by main grants from the Netherlands Organisation for Scientific Research (GB-MAGW 480-03-005, GB-MAGW 480-08-006), Stichting Achmea Slachtoffer en Samenleving (SASS), the Netherlands Organisation for Scientific Research to the Consortium Individual Development (CID; 024.001.003), a grant of the European Research Council (ERC-2017-CoG - 773023 INTRANSITION), and various other grants from the Netherlands Organisation for Scientific Research, VU University Amsterdam, and Utrecht University. The funders had no role in the study design, data collection and analysis, decision to publish, or preparation of the manuscript. If the regulations of the journal allow it, the participating children, parents and schools are to be thanked as well.

abstract: |
  This paper set out to identify the most important early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM forests. Candidate predictors were selected based on a prior text mining systematic review of 6305 papers. Participants were 497 Dutch adolescents. Predictors were assessed at age 13, and emotion regulation development from age 14-18 was described using a quadratic latent trajectory model. SEM-forests were used to group adolescents based on their predictor scores, while maximizing the similarity of developmental trajectories within each group. Compared to other person-centered approaches, this method not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents follow similar trajectories. The results indicated that the most important early predictors (at age 13) were individual differences (e.g., in personality), conflict behaviors towards parents and friends, and internalizing and externalizing behavior. Less important were parenting behaviors, except autonomy support and invasive parenting, bullying and victimization, delinquent peers, and substance use. The implications of these findings for theory formation and intervention are discussed, and we present an open source risk assessment tool, ERRATA.

@@

Emotion regulation,
defined as the ability to manage emotional experience and expression,
is critical for mental health [@aldaoEmotionregulationStrategiesPsychopathology2010, @schaferEmotionRegulationStrategies2017, @braetEmotionRegulationChildren2014]
and social competence [@zimmermannEmotionRegulationEarly2014] throughout life.
There is mounting evidence, however, that adolescents do not uniformly become better at regulating emotions: 
staggered development of emotional and regulatory brain circuits gives rise to a "maturity gap" in mid-adolescence [@croneUnderstandingAdolescencePeriod2012].
The emotional salience of social stimuli is intensified, while the ability to regulate these emotions lags behind [@caseyBrakingAcceleratingAdolescent2011, @vanlissaLongitudinalInterplayAffective2014].
This imbalance motivates most adolescents to engage in adaptive exploration:
Seeking out experiences required to develop mature socio-emotional competences.
At the same time, this imbalance renders some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014].
<!-- exposes youth to a risk of emotion dysregulation, -->
For example, because they are not yet well-equipped to regulate the pursuit of unhealthy incentives,
or persevere in pursuing adaptive goals when facing obstacles.
Thus, some become trapped in a negative spiral characterized by more negative and volatile emotions [@craccoEmotionRegulationChildhood2017],
which has been linked to the emergence of several mental disorders [@schaferEmotionRegulationStrategies2017].
<!-- This imbalance motivates most youngsters to seek out new challenges in life and love, -->
<!-- but renders some of them vulnerable to emotion dysregulation1,10. Unfortunately, there is little theory about which risk factors and environmental hazards render adolescents susceptible to emotional difficulties11. -->
<!-- Acquiring adaptive emotion regulation skills in adolescence plays a key role in this regard7,20, because t -->

@@

As we set out to identifying important predictors of emotion regulation development,
this latent class analysis is not sufficient to answer our research question.
In latent class analysis, individuals are clustered only with respect the the outcome (growth trajectories), to ensure maximum homogeneity of growth trajectories in each class.
It thus answers the question: "Which adolescents have similar growth trajectories"?
SEM-forests similarly aim to ensure maximum homogeneity of growth trajectories within clusters,
but they differ from latent class analysis in that they cluster individuals by reference to a set of predictors.
They thus answer the question: "Why do some adolescents have similar growth trajectories"?
This method is suitable to meet the study goals.
